---
title: "Project proposal"
author: "Team MBK"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(gapminder)
library(broom)
library(skimr)
```

## 1. Introduction

The introduction should introduce your general research question(s) and your data (where it came from, how it was collected, what are the cases, what are the variables, etc.).

Our project will look at how social and economic patterns have changed in different countries overtime. We plan to use the Gapminder dataset, which tracks social and economic indicators (like population, GDP per Capita, and Fertility overtime. 

Our guiding questions are:
- How are GDP per capita and life expectancy related across countries?
- Have these relationships changed by country over time?

Understanding these questions helps reveal how economic growth influences population well-being, and whether global inequality in health and wealth is narrowing or widening.

The Gapminder dataset was compiled by Hans Rosling’s Gapminder Foundation, using publicly available data from the World Bank, United Nations, and OECD. Each row in the dataset represents a country in a given year, with variables describing life expectancy, population, and GDP per capita.

## 2. Data

Text goes here. Place your data in the /data folder, and add dimensions and codebook to the README in that folder. Then print out the output of glimpse() or skim() of your data frame.

```{r}
library(readr)
homicide <- read_csv("../data/murder_total_deaths.csv")
life <- read_csv("../data/lex.csv")
gdp_per_capita <- read_csv("../data/gdp_pcap.csv")
fertility <- read_csv("../data/children_per_woman_total_fertility.csv")
population <- read_csv("../data/pop.csv")
```

## 3. Data analysis plan

Text goes here.
- What variables will you visualize to explore your research questions?
- Will there be any other data that you need to find to help with your research question?
- Very preliminary exploratory data analysis, including some summary statistics and visualizations, along with some explanation on how they help you learn more about your data. (You can add to these later as you work on your project.)
- The data visualization(s) that you believe will be useful in exploring your question(s). (You can update these later as you work on your project.)


```{r}
# Code goes here
# Code to calculate summary statistics
# Code for a visualization
```
