---
title: "Project memo"
author: "Team MBK"
output: github_document
---


This document contains a detailed account of the data cleanup steps and design choices for our visualizations exploring the relationship between GDP per capita and social outcomes (homicide rates, life expectancy, and fertility) across countries from 1990-2021.
```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(readr)
library(scales)
library(dplyr)
library(ggplot2)
```

## Data Clean Up Steps for Overall Data

### Step 1: Load and Pivot Individual Datasets to Long Format

We load five separate Gapminder datasets and convert each from wide format (years as columns) to long format (year as a variable) for easier analysis and merging.
```{r load-and-pivot-data, message = FALSE}
homicide <- read_csv("../data/murder_total_deaths.csv")
life <- read_csv("../data/lex.csv")
gdp_per_capita <- read_csv("../data/gdp_pcap.csv")
babies_per_woman <- read_csv("../data/children_per_woman_total_fertility.csv")
population <- read_csv("../data/pop.csv")

homicide_long <- homicide %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "homicide_rate"
  )

life_long <- life %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "lifeExp"
  ) 

gdp_long <- gdp_per_capita %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "gdpPercap"
  ) 

babies_per_woman_long <- babies_per_woman %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "babies_per_woman"
  ) 

pop_long <- population %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "pop"
  )
```

### Step 2: Merge Datasets and Filter Years

We merge all five datasets using full joins to preserve all country-year observations, then filter to focus on 1990-2021. We also calculate growth rates for GDP and population.
```{r merge-and-filter-data}
combined_long <- homicide_long %>%
  full_join(life_long, by = c("geo", "name", "year")) %>%
  full_join(gdp_long, by = c("geo", "name", "year")) %>%
  full_join(babies_per_woman_long, by = c("geo", "name", "year")) %>%
  full_join(pop_long, by = c("geo", "name", "year"))

combined_long <- combined_long %>%
  filter(year >= 1990 & year <= 2021)

combined_long <- combined_long %>%
  mutate(year = as.numeric(year)) %>%
  group_by(geo) %>% 
  mutate(
    gdp_growth_percent = (gdpPercap - lag(gdpPercap)) / lag(gdpPercap) * 100,
    pop_growth_percent = (pop - lag(pop)) / lag(pop) * 100
  ) %>%
  ungroup()
```

### Step 3: Create Income Group Classification

We classify countries into four income groups based on GDP per capita thresholds from the World Bank. This classification will help us visualize patterns across different economic development levels within each of our graphs.
```{r create-income-groups}
combined_long <- combined_long %>%
  mutate(
    income_group = case_when(
      gdpPercap >= 50000 ~ "High GDP",
      gdpPercap >= 15000 ~ "Upper-Middle GDP",
      gdpPercap >= 4000 ~ "Lower-Middle GDP",
      TRUE ~ "Low GDP"
    ),
    income_group = factor(
      income_group, 
      levels = c("Low GDP", "Lower-Middle GDP", "Upper-Middle GDP", "High GDP")
    )
  )
```

## Color Palette

We use a colorblind-friendly palette throughout our visualizations. The colors are distinguishable for people with various types of color vision deficiency and are also labeled in legends for double-coding accessibility.
```{r define-color-palette}
income_colors <- c(
  "High GDP" = "#2A9D8F",
  "Upper-Middle GDP" = "#E9C46A",
  "Lower-Middle GDP" = "#F4A261",
  "Low GDP" = "#E76F51"
)
```

---

## Plots
### Plot 1: Murder Rate x GDP Per Capita

```{r plot-murder-rate, fig.alt = "Scatter plot showing GDP per capita (log scale) on the x-axis and homicide rate per 100,000 people on the y-axis for countries in 2020. Points are colored by income group: Low GDP (coral), Lower-Middle GDP (orange), Upper-Middle GDP (yellow), and High GDP (teal). A dashed gray linear trend line shows a negative relationship, indicating that countries with higher GDP per capita tend to have lower homicide rates."}

plot_data <- combined_long %>%
  filter(year == 2020,
         !is.na(gdpPercap),
         !is.na(homicide_rate))

murder_rate_plot <- ggplot(plot_data, aes(x = gdpPercap, y = homicide_rate)) +
  geom_point(aes(color = income_group)) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    linetype = "dashed",
    linewidth = 1.1,
    color = "grey40"
  ) +
  scale_x_log10() +
  scale_color_manual(
    values = income_colors,
    drop = FALSE,
    name = "Income Group"
  ) +
  labs(
    title = "GDP Per Capita vs Murder Rate",
    subtitle = "Year: 2020",
    x = "GDP per Capita (log scale)",
    y = "Homicide Rate (per 100k)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.title = element_blank(),
    aspect.ratio = .75
  )

murder_rate_plot

ggsave("murder_rate_X_GDP_per_capita_2020.png")
```


### Plot 2: Life Expectancy x GDP Per Capita


```{r plot-life-expectancy, fig.alt = "Scatter plot showing GDP per capita (log scale) on the x-axis and life expectancy in years on the y-axis for countries in 2020. Points are colored by income group: Low GDP (coral), Lower-Middle GDP (orange), Upper-Middle GDP (yellow), and High GDP (teal). A dashed gray linear trend line shows a positive relationship, indicating that countries with higher GDP per capita tend to have higher life expectancy."}

plot_data <- combined_long %>%   
  filter(year == 2020,
         !is.na(gdpPercap),
         !is.na(lifeExp))

gdp_life_plot <- ggplot(plot_data, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(color = income_group)) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    linetype = "dashed",
    linewidth = 1.1,
    color = "grey40"
  ) +
  scale_x_log10() +
  scale_color_manual(
    values = income_colors,
    drop = FALSE,
    name = "Income Group"
  ) +
  labs(
    title = "GDP vs Life Expectancy",
    subtitle = "Year: 2020",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy (years)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.title = element_blank(),
    aspect.ratio = .75
)

gdp_life_plot

ggsave("life_expectancy_X_GDP_per_capita_2020.png")

```


### Plot 3: Fertility Rate x GDP Per Capita 

```{r plot-fertility-rate, fig.alt = "Scatter plot showing GDP per capita (log scale) on the x-axis and fertility rate (babies per woman) on the y-axis for countries in 2020. Points are colored by income group: Low GDP (coral), Lower-Middle GDP (orange), Upper-Middle GDP (yellow), and High GDP (teal). A dashed gray linear trend line shows a negative relationship, indicating that countries with higher GDP per capita tend to have lower fertility rates."}

plot_data <- combined_long %>%
  filter(year == 2020,
         !is.na(gdpPercap),
         !is.na(babies_per_woman))

fertility_rate_plot <- ggplot(plot_data, aes(x = gdpPercap, y = babies_per_woman)) +
  geom_point(aes(color = income_group)) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    linetype = "dashed",
    linewidth = 1.1,
    color = "grey40"
  ) +
  scale_x_log10() +
  scale_color_manual(
    values = income_colors,
    drop = FALSE,
    name = "Income Group"
  ) +
  labs(
    title = "GDP Per Capita vs Fertility Rate",
    subtitle = "Year: 2020",
    x = "GDP per Capita (log scale)",
    y = "Fertility Rate (Babies per Woman)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.title = element_blank(),
    aspect.ratio = .75
)

fertility_rate_plot

ggsave("fertility_rate_X_GDP_per_capita_2020.png")
```

### Plot 4: Population x GDP Per Capita 

```{r plot-population, fig.alt = "Scatter plot showing GDP per capita (log scale) on the x-axis and population size on the y-axis for countries in 2020. Points are colored by income group: Low GDP (coral), Lower-Middle GDP (orange), Upper-Middle GDP (yellow), and High GDP (teal). A dashed gray linear trend line is included but shows a weak or unclear relationship, indicating that population size does not have a strong linear association with GDP per capita."}
plot_data <- combined_long %>%
  filter(year == 2020,
         !is.na(gdpPercap),
         !is.na(pop))

population_plot <- ggplot(plot_data, aes(x = gdpPercap, y = pop)) +
  geom_point(aes(color = income_group)) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    linetype = "dashed",
    linewidth = 1.1,
    color = "grey40"
  ) +
  scale_x_log10() +
  scale_color_manual(
    values = income_colors,
    drop = FALSE,
    name = "Income Group"
  ) +
  labs(
    title = "GDP Per Capita vs Population Size",
    subtitle = "Year: 2020",
    x = "GDP per Capita (log scale)",
    y = "Population Size"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    legend.title = element_blank(),
    aspect.ratio = .75
)

population_plot

ggsave("population_X_GDP_per_capita_2020.png")
```
```


### Calculating differences
```{r}
# Wealthiest Country
wealthiest <- combined_long %>%
  filter(year == 2020) %>%
  slice_max(gdpPercap, n = 1) %>%
  select(name, gdpPercap, lifeExp, pop, babies_per_woman, homicide_rate)

#Poorest Country
poorest <- combined_long %>%
  filter(year == 2020) %>%
  slice_min(gdpPercap, n = 1) %>%
  select(name, gdpPercap, lifeExp, pop, babies_per_woman, homicide_rate)

#GDP Difference Between Wealthiest and Poorest Countries
gdp_difference <- wealthiest$gdpPercap - poorest$gdpPercap
gdp_difference

#Life Expectancy Difference
life_expectancy_difference <- wealthiest$lifeExp - poorest$lifeExp
life_expectancy_difference

#Population Difference
pop_difference <- wealthiest$pop - poorest$pop
pop_difference

#Fertility Difference
fertility_difference <- wealthiest$babies_per_woman - poorest$babies_per_woman
fertility_difference

```
```