---
title: "Project memo"
author: "Team MBK"
output: github_document
---

This document should contain a detailed account of the data clean up for your data and the design choices you are making for your plots. For instance you will want to document choices you've made that were intentional for your graphic, e.g. color you've chosen for the plot. Think of this document as a code script someone can follow to reproduce the data cleaning steps and graphics in your handout.

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(readr)
library(scales)
```

## Data Clean Up Steps for Overall Data

### Step 1: Load and Pivot Individual Datasets to Long Format
We load five separate CSV files from Gapminder and pivot each from wide format (years as columns) to long format (one row per country-year). This structure is necessary for merging datasets and performing analysis.
```{r}
homicide <- read_csv("../data/murder_total_deaths.csv")
life <- read_csv("../data/lex.csv")
gdp_per_capita <- read_csv("../data/gdp_pcap.csv")
babies_per_woman <- read_csv("../data/children_per_woman_total_fertility.csv")
population <- read_csv("../data/pop.csv")

homicide_long <- homicide %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "homicide_rate"
  )

life_long <- life %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "lifeExp"
  ) 

gdp_long <- gdp_per_capita %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "gdpPercap"
  ) 

babies_per_woman_long <- babies_per_woman %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "babies_per_woman"
  ) 

pop_long <- population %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "pop"
  )
```

### Step 2: Merge Datasets and Filter Years
We merge all five datasets using full joins to preserve all country-year observations, then filter to focus on 1990-2021. We also calculate growth rates for GDP and population.
```{r}
combined_long <- homicide_long %>%
  full_join(life_long, by = c("geo", "name", "year")) %>%
  full_join(gdp_long, by = c("geo", "name", "year")) %>%
  full_join(babies_per_woman_long, by = c("geo", "name", "year")) %>%
  full_join(pop_long, by = c("geo", "name", "year"))

combined_long <- combined_long %>%
  filter(year >= 1990 & year <= 2021)

combined_long <- combined_long %>%
  mutate(year = as.numeric(year)) %>%
  group_by(geo) %>% 
  mutate(
    gdp_growth_percent = (gdpPercap - lag(gdpPercap)) / lag(gdpPercap) * 100,
    pop_growth_percent = (pop - lag(pop)) / lag(pop) * 100
  ) %>%
  ungroup()

```

## Plots

### ggsave example for saving plots

```{r starwars ggsave example}
p1 <- starwars |>
  filter(mass < 1000, 
         species %in% c("Human", "Cerean", "Pau'an", "Droid", "Gungan")) |>
  ggplot() +
  geom_point(aes(x = mass, 
                 y = height, 
                 color = species)) +
  labs(x = "Weight (kg)", 
       y = "Height (m)",
       color = "Species",
       title = "Weight and Height of Select Starwars Species",
       caption = paste("This data comes from the starwars api: https://swapi.py43.com"))


ggsave("example-starwars.png", width = 4, height = 4)

ggsave("example-starwars-wide.png", width = 6, height = 4)
```

# Plot #1: Life Expectancy Comparison - Top 3 vs Bottom 3 GDP Countries

This bar chart provides a direct comparison of life expectancy between the world's wealthiest and poorest countries in 2021, clearly illustrating the health gap associated with economic development.

Design choices:
- **2021 focus**: Uses the most recent year in the dataset for current relevance
- **Top 3 vs Bottom 3 comparison**: Highlights the extremes to show the maximum health disparity between richest and poorest nations
- **Horizontal bars (coord_flip)**: Country names are easier to read horizontally, and bars naturally show magnitude
- **Ordered by life expectancy**: Countries arranged from lowest to highest life expectancy for intuitive reading
- **Color contrast**: Dark green for wealthy countries and coral for poor countries creates clear visual separation between groups
- **Legend at bottom**: Keeps the focus on the bars while maintaining clarity about what each color represents
- **Centered bold title**: Emphasizes the comparison being made
- **Minimal theme**: Clean, professional appearance without distracting elements

# Plot #1 Final
```{r}
plot1_data <- combined_long %>%
  filter(!is.na(gdpPercap), !is.na(lifeExp), year == 2021)

# Find top 3 and bottom 3 countries by GDP per capita
top3 <- plot1_data %>%
  arrange(desc(gdpPercap)) %>%
  slice_head(n = 3) %>%
  mutate(group = "Top 3 GDP")

bottom3 <- plot1_data %>%
  arrange(gdpPercap) %>%
  slice_head(n = 3) %>%
  mutate(group = "Bottom 3 GDP")

# Combine the two sets
compare_data <- bind_rows(top3, bottom3)

# Plot bar chart
p1 <- compare_data %>%
  ggplot(aes(x = reorder(name, lifeExp), y = lifeExp, fill = group)) +
  geom_col(width = 0.7) +
  coord_flip() +
  labs(
    title = "Life Expectancy: Top 3 vs Bottom 3 GDP per Capita Countries (2021)",
    x = "Country",
    y = "Life Expectancy (years)",
    fill = "Group",
    caption = "Data source: Gapminder (World Bank)"
  ) +
  scale_fill_manual(values = c("Top 3 GDP" = "darkgreen", "Bottom 3 GDP" = "coral")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
    legend.position = "bottom"
  )

p1

ggsave("plot_top_bottom3_gdp_life_expectancy.png", plot = p1, width = 8, height = 5)

```
### Plot 2: _________

### Plot 3: ___________

Add more plot sections as needed. Each project should have at least 3 plots, but talk to me if you have fewer than 3.

### Plot 4: ___________ 

