---
title: "Project memo"
author: "Team name"
output: github_document
---


This document contains a detailed account of the data cleanup steps and design choices for our visualizations exploring the relationship between GDP per capita and social outcomes (homicide rates, life expectancy, and fertility) across countries from 1990-2021.
```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(readr)
library(scales)
```

## Data Clean Up Steps for Overall Data

### Step 1: Load and Pivot Individual Datasets to Long Format

We load five separate Gapminder datasets and convert each from wide format (years as columns) to long format (year as a variable) for easier analysis and merging.
```{r load-and-pivot-data, message = FALSE}
homicide <- read_csv("../data/murder_total_deaths.csv")
life <- read_csv("../data/lex.csv")
gdp_per_capita <- read_csv("../data/gdp_pcap.csv")
babies_per_woman <- read_csv("../data/children_per_woman_total_fertility.csv")
population <- read_csv("../data/pop.csv")

homicide_long <- homicide %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "homicide_rate"
  )

life_long <- life %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "lifeExp"
  ) 

gdp_long <- gdp_per_capita %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "gdpPercap"
  ) 

babies_per_woman_long <- babies_per_woman %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "babies_per_woman"
  ) 

pop_long <- population %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "pop"
  )
```

### Step 2: Merge Datasets and Filter Years

We merge all five datasets using full joins to preserve all country-year observations, then filter to focus on 1990-2021. We also calculate growth rates for GDP and population.
```{r merge-and-filter-data}
combined_long <- homicide_long %>%
  full_join(life_long, by = c("geo", "name", "year")) %>%
  full_join(gdp_long, by = c("geo", "name", "year")) %>%
  full_join(babies_per_woman_long, by = c("geo", "name", "year")) %>%
  full_join(pop_long, by = c("geo", "name", "year"))

combined_long <- combined_long %>%
  filter(year >= 1990 & year <= 2021)

combined_long <- combined_long %>%
  mutate(year = as.numeric(year)) %>%
  group_by(geo) %>% 
  mutate(
    gdp_growth_percent = (gdpPercap - lag(gdpPercap)) / lag(gdpPercap) * 100,
    pop_growth_percent = (pop - lag(pop)) / lag(pop) * 100
  ) %>%
  ungroup()
```

### Step 3: Create Income Group Classification

We classify countries into four income groups based on GDP per capita thresholds from the World Bank. This classification helps visualize patterns across different economic development levels.
```{r create-income-groups}
combined_long <- combined_long %>%
  mutate(
    income_group = case_when(
      gdpPercap >= 50000 ~ "High Income",
      gdpPercap >= 15000 ~ "Upper-Middle Income",
      gdpPercap >= 4000 ~ "Lower-Middle Income",
      TRUE ~ "Low Income"
    ),
    income_group = factor(
      income_group, 
      levels = c("Low Income", "Lower-Middle Income", "Upper-Middle Income", "High Income")
    )
  )
```

## Color Palette

We use a colorblind-friendly palette throughout our visualizations. The colors are distinguishable for people with various types of color vision deficiency and are also labeled in legends for double-coding accessibility.
```{r define-color-palette}
income_colors <- c(
  "High Income" = "#2A9D8F",
  "Upper-Middle Income" = "#E9C46A",
  "Lower-Middle Income" = "#F4A261",
  "Low Income" = "#E76F51"
)
```

---

## Plots
### Plot 1: Life Expectancy Comparison - Top 3 vs Bottom 3 GDP Countries

This bar chart provides a direct comparison of life expectancy between the world's wealthiest and poorest countries in 2021, clearly illustrating the health gap associated with economic development.

Design choices:
- **2021 focus**: Uses the most recent year in the dataset for current relevance
- **Top 3 vs Bottom 3 comparison**: Highlights the extremes to show the maximum health disparity between richest and poorest nations
- **Horizontal bars (coord_flip)**: Country names are easier to read horizontally, and bars naturally show magnitude
- **Ordered by life expectancy**: Countries arranged from lowest to highest life expectancy for intuitive reading
- **Color contrast**: Dark green for wealthy countries and coral for poor countries creates clear visual separation between groups
- **Legend at bottom**: Keeps the focus on the bars while maintaining clarity about what each color represents
- **Centered bold title**: Emphasizes the comparison being made
- **Minimal theme**: Clean, professional appearance without distracting elements

#### Data cleanup steps specific to plot 1

Filter to 2021 data only, then identify the top 3 and bottom 3 countries by GDP per capita.
```{r}
plot1_data <- combined_long %>%
  filter(!is.na(gdpPercap), !is.na(lifeExp), year == 2021)

# Find top 3 and bottom 3 countries by GDP per capita
top3 <- plot1_data %>%
  arrange(desc(gdpPercap)) %>%
  slice_head(n = 3) %>%
  mutate(group = "Top 3 GDP")

bottom3 <- plot1_data %>%
  arrange(gdpPercap) %>%
  slice_head(n = 3) %>%
  mutate(group = "Bottom 3 GDP")

# Combine the two sets
compare_data <- bind_rows(top3, bottom3)
```

#### Final Plot 1
```{r}
p1 <- compare_data %>%
  ggplot(aes(x = reorder(name, lifeExp), y = lifeExp, fill = group)) +
  geom_col(width = 0.7) +
  coord_flip() +
  labs(
    title = "Life Expectancy: Top 3 vs Bottom 3 GDP per Capita Countries (2021)",
    x = "Country",
    y = "Life Expectancy (years)",
    fill = "Group",
    caption = "Data source: Gapminder (World Bank)"
  ) +
  scale_fill_manual(values = c("Top 3 GDP" = "darkgreen", "Bottom 3 GDP" = "coral")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
    legend.position = "bottom"
  )

p1


```


### Plot 2: _________


```{r}
library(dplyr)
library(ggplot2)
data_2020 <- combined_long %>%
  filter(year == 2020, !is.na(gdpPercap), !is.na(lifeExp))
plot_data <- data_2020 %>%
  mutate(
    gdp_level = ntile(gdpPercap, 3),       
    gdp_level = factor(
      gdp_level,
      levels = c(1, 2, 3),
      labels = c("Low GDP", "Middle GDP", "High GDP")
    )
  )
p1 <- ggplot(plot_data, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(color = gdp_level)) +
geom_smooth(
    method = "lm",
    se = FALSE,
    linetype = "dashed",
    color = "grey20"
  ) +
  scale_x_log10() +
  scale_color_manual(
    values = c(
      "Low GDP"    = "#F4A582",
      "Middle GDP" = "#FDBF6F",
      "High GDP"   = "#7FC97F"
    )
  ) +
  labs(
    title   = "GDP vs Life Expectancy",
    x       = "GDP per Capita (log scale)",
    y       = "Life Expectancy (years)",
    color   = "GDP Level"
  ) +
  theme_minimal() 
p1
ggsave("GDP-vs-Life-Expectancy-scatter.png", width = 8, height = 5)

```


### Plot 3: ___________

Add more plot sections as needed. Each project should have at least 3 plots, but talk to me if you have fewer than 3.

### Plot 4: ___________ 

