---
title: "Project memo"
author: "Team name"
output: github_document
---

This document should contain a detailed account of the data clean up for your data and the design choices you are making for your plots. For instance you will want to document choices you've made that were intentional for your graphic, e.g. color you've chosen for the plot. Think of this document as a code script someone can follow to reproduce the data cleaning steps and graphics in your handout.

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(readr)
library(scales)
```

## Data Clean Up Steps for Overall Data

### Step 1: Load and Pivot Individual Datasets to Long Format

```{r}
homicide <- read_csv("../data/murder_total_deaths.csv")
life <- read_csv("../data/lex.csv")
gdp_per_capita <- read_csv("../data/gdp_pcap.csv")
babies_per_woman <- read_csv("../data/children_per_woman_total_fertility.csv")
population <- read_csv("../data/pop.csv")

homicide_long <- homicide %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "homicide_rate"
  )

life_long <- life %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "lifeExp"
  ) 

gdp_long <- gdp_per_capita %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "gdpPercap"
  ) 

babies_per_woman_long <- babies_per_woman %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "babies_per_woman"
  ) 

pop_long <- population %>%
  pivot_longer(
    cols = -c(geo, name),
    names_to  = "year",
    values_to = "pop"
  )
```

### Step 2: Merge Datasets and Filter Years
We merge all five datasets using full joins to preserve all country-year observations, then filter to focus on 1990-2021. We also calculate growth rates for GDP and population.
```{r}
combined_long <- homicide_long %>%
  full_join(life_long, by = c("geo", "name", "year")) %>%
  full_join(gdp_long, by = c("geo", "name", "year")) %>%
  full_join(babies_per_woman_long, by = c("geo", "name", "year")) %>%
  full_join(pop_long, by = c("geo", "name", "year"))

combined_long <- combined_long %>%
  filter(year >= 1990 & year <= 2021)

combined_long <- combined_long %>%
  mutate(year = as.numeric(year)) %>%
  group_by(geo) %>% 
  mutate(
    gdp_growth_percent = (gdpPercap - lag(gdpPercap)) / lag(gdpPercap) * 100,
    pop_growth_percent = (pop - lag(pop)) / lag(pop) * 100
  ) %>%
  ungroup()

```

## Plots

### ggsave example for saving plots

```{r starwars ggsave example}
p1 <- starwars |>
  filter(mass < 1000, 
         species %in% c("Human", "Cerean", "Pau'an", "Droid", "Gungan")) |>
  ggplot() +
  geom_point(aes(x = mass, 
                 y = height, 
                 color = species)) +
  labs(x = "Weight (kg)", 
       y = "Height (m)",
       color = "Species",
       title = "Weight and Height of Select Starwars Species",
       caption = paste("This data comes from the starwars api: https://swapi.py43.com"))


ggsave("example-starwars.png", width = 4, height = 4)

ggsave("example-starwars-wide.png", width = 6, height = 4)
```

### Step 2: ________

## Plots

### ggsave example for saving plots

```{r starwars ggsave example}
p1 <- starwars |>
  filter(mass < 1000, 
         species %in% c("Human", "Cerean", "Pau'an", "Droid", "Gungan")) |>
  ggplot() +
  geom_point(aes(x = mass, 
                 y = height, 
                 color = species)) +
  labs(x = "Weight (kg)", 
       y = "Height (m)",
       color = "Species",
       title = "Weight and Height of Select Starwars Species",
       caption = paste("This data comes from the starwars api: https://swapi.py43.com"))


ggsave("example-starwars.png", width = 4, height = 4)

ggsave("example-starwars-wide.png", width = 6, height = 4)
```



### Plot 1: _________

#### Data cleanup steps specific to plot 1

This plot examines the relationship between economic prosperity (GDP per capita) and health outcomes (life expectancy).

Design choices:
- Log scale on x-axis to handle the wide range of GDP values and reveal patterns across all income levels
- Steelblue color with alpha = 0.4 transparency to show density where points overlap
- LOESS smoothing curve in dark red to highlight the overall positive relationship
- Dollar formatting on x-axis for readability
- Minimal theme for clean, professional appearance

```{r}
plot1_data <- combined_long %>%
  filter(!is.na(gdpPercap), !is.na(lifeExp))
```


#### Final Plot 1

```{r}
p1 <- plot1_data %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point(alpha = 0.4, color = "steelblue", size = 1.5) +
  geom_smooth(method = "loess", color = "darkred", se = TRUE, linewidth = 1.2) +
  scale_x_log10(labels = dollar_format()) +
  labs(
    title = "GDP per Capita and Life Expectancy",
    subtitle = "All countries, 1990-2021",
    x = "GDP per Capita (log scale, USD)",
    y = "Life Expectancy (years)",
    caption = "Data source: Gapminder (World Bank)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "gray40")
  )
p1
ggsave("plot1_gdp_life_expectancy.png", plot = p1, width = 7, height = 5)
```


### Plot 2: _________


### Plot 3: ___________

Add more plot sections as needed. Each project should have at least 3 plots, but talk to me if you have fewer than 3.

### Plot 4: ___________ 

